Exploring probability distributions for bivariate temporal granularities

### 2019-12-09 main.Rmd
changes till quasi-circular

begin{equation}
pseudomonth & = Alter^{12}_{11,-1}(day, \\
  Alter^{12}_{9,-1}(day, \\
  Alter^{12}_{6,-1}(day,  \\
  Alter^{12}_{4,-1)}(day, \\
  Alter^{12}_{(2,-3)}(day, \\
  Group_{31}(day))))))
end{equation}

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A time domain, as defined by [@Bettini1998-ed], is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. Lining up with the the assumption in [@Bettini2000-vy] that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists and is represented by the index of the tsibble. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

Cyclic time granularities which accommodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. To elaborate more, mappings will be regular if, for example, linear granularities are formed by grouping operation. Grouping operations combine fixed granules of a linear granularity to form a new granule of the resulting linear granularity. However, if granularities are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of the root granularity. Cyclic time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the same is irregular. Examples of circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month. The formalism would differ for circular and quasi-circular granularities although both accommodate periodicities in data.

@@

\begin{definition}\label{def:circular}
A circular granularity $C_{B, G}$ relates a linear granularity $G$ to the bottom granularity B, if

@@

**Example**:  Example showing circular granularity relating two linear granularities "Day" and "Week" is visually depicted in a series of slots in the \autoref{fig:circular-dow}. Each granule is represented by a box. The diagram also illustrates that the granules that overlap share elements from the underlying time domain. The first slot in the diagram shows the index set. The index of the tsibble considered is days. Day and Week are two linear granularities with $P =  7$. The circular granularity $C_{Day, Week}$ representing Day-of-Week will thus consist of pattern {L(0), L(1), L(2), L(3), L(4), L(5), L(6)} which repeats itself after each period length.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\toprule
Minute-of-Hour: & $C_1$ & $= L_1(z \mod 60)$ & $P_1 =$&$60$ \\
Minute-of-Day: & $C_2$ & $= L_2(z \mod 60*24)$ & $P_2=$&$`r 60*24`$\\
Hour-of-Day: & $C_3$ & $= L_3(\lfloor z/60\rfloor\mod 24)$  & $P_3 =$&$24$ \\
Hour-of-Week: & $C_4$ & $= L_4(\lfloor z/60\rfloor\mod 24*7)$  & $P_4=$&$`r 24*7`$\\
Day-of-Week: & $C_5$ & $= L_5(\lfloor z/24*60\rfloor \mod 7)$ & $P_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative circular granularities with time index in minutes}
\label{tab:definitions}
\end{table}

@@

A **quasi-circular** granularity can not be defined using modular arithmetic due to its irregular mapping with the bottom granularity. However, they are still formed with linear granularities, one of which "groups periodically into" the other. \autoref{tab:quasi} shows some example of quasi-circular granularities ($Q_i$) with ($P_i$) denoting the plausible choices of number of granules of the finer granularity inside each granule of the coarser one.

@@

\begin{definition}\label{def:quasicircular}
A quasi-circular granularity $Q_{B, G'}$ that relates linear granularities $G'$ and bottom granularity $B$, if

>>>>

\begin{equation}
\begin{split}
pseudomonth & = Alter^{12}_{11,-1}(day, \\
  Alter^{12}_{9,-1}(day, \\
  Alter^{12}_{6,-1}(day,  \\
  Alter^{12}_{4,-1)}(day, \\
  Alter^{12}_{(2,-3)}(day, \\
  Group_{31}(day))))))
\end{split}
\end{equation}

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and key is a set of variables that define observational units over time. In a tsibble each observation should be uniquely identified by index and key.

A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble. Lining up with the the assumption in @Bettini2000-vy, that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

Cyclic time granularities which accommodate for periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might have days ranging from 28 to 31. To elaborate more, mappings will be regular if, for example, linear granularities are formed by grouping operation. Grouping operations combine fixed granules of a linear granularity to form a new granule of the resulting linear granularity. However, if granularities are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of the root granularity. Cyclic time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the same is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month. The formalism would differ for circular and quasi-circular granularities although both accommodate for periodicities in data.

@@

\begin{definition}\label{def:circular}
(circular granularity): A circular granularity $C_{B, G}$ relates a linear granularity $G$ to the bottom granularity B, if

@@

**Example**:  Example showing circular granularity relating two linear granularities "day" and "week" is visually depicted in a series of slots in the \autoref{fig:circular-dow}. Each granule is represented by a box. The diagram also illustrates that the granules that overlap share elements from the underlying time domain. The first slot in the diagram shows the index set. The bottom granularity, represented by the index of the tsibble considered is "day". "day" and "week" are two linear granularities with $P =  7$. The circular granularity $C_{day, week}$ representing day-of-week will thus consist of pattern {L(0), L(1), L(2), L(3), L(4), L(5), L(6)} which repeats itself after each period length.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\toprule
minute-of-Hour: & $C_1$ & $= L_1(z \mod 60)$ & $P_1 =$&$60$ \\
minute-of-Day: & $C_2$ & $= L_2(z \mod 60*24)$ & $P_2=$&$`r 60*24`$\\
hour-of-Day: & $C_3$ & $= L_3(\lfloor z/60\rfloor\mod 24)$  & $P_3 =$&$24$ \\
hour-of-Week: & $C_4$ & $= L_4(\lfloor z/60\rfloor\mod 24*7)$  & $P_4=$&$`r 24*7`$\\
day-of-Week: & $C_5$ & $= L_5(\lfloor z/24*60\rfloor \mod 7)$ & $P_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative circular granularities with time index in minutes}
\label{tab:definitions}
\end{table}

@@

A **quasi-circular** granularity can not be defined using modular arithmetic since they are formed using two linear granularities with irregular mapping. However, they are still formed with linear granularities, one of which "groups periodically into" the other. \autoref{tab:quasi} shows some example of quasi-circular granularities ($Q_i$) with ($P_i$) denoting the plausible choices of number of granules of the finer granularity inside each granule of the coarser one.

@@

\begin{definition}\label{def:quasicircular}
(quasi-circular granularity): A quasi-circular granularity $Q_{B, G'}$ that relates linear granularities $G'$ and bottom granularity $B$, if
